### 2025-10-26 thesis/implementation/index.tex
Write about voting

Similar to ICL generation, refined responses are processed through the same
\texttt{Markdown} parser to extract the predicted SQL. The system validates each
extracted candidate using \texttt{QueryCandidate::try_from} and returns the first
syntactically valid refinement starting from the bottom. This ensures that
refinement produces executable SQL whilst falling back gracefully if refinement
fails.

@@

>>>>

Similar to ICL generation, refined responses are processed through the same
\texttt{Markdown} parser to extract the predicted SQL. The system validates each
extracted candidate by trying to parse it using \texttt{QueryCandidate::try\_from}
and returns the first syntactically valid refinement starting from the bottom.
This ensures that refinement produces executable SQL whilst falling back
gracefully if refinement fails.

@@

\subsubsection{Consensus Voting System ($\nu$)}

The \texttt{natural-inference::pipeline::voting} implements the majority
function $\nu$ – this implementation closely follows the design and algorithm
outlined in the section~\ref{design:voting-function}. The function \texttt{voting}
providing a result-based self-consensus mechanism which takes in the set of
candidates that were predicted during the generation phase and returns the most
likely query.

\paragraph{Result-Based Self-Consensus}

\citeauthor{OmniSQL} have shown in \citeyear{OmniSQL} that self-consensus can 
significantly improve the accuracy of models that are already showing 
state-of-the-art performance. The result-based self-consensus mechanism is
executing every query candidate, verifies that it works on the database, and
loads it's results. By partitioning the available candidates into buckets based
on their hashed result, queries can be deemed semantically equivalent if they
end up in the same bucket. After every query has been exeuted and partitioned,
the algorithm groups the buckets in a hash map \texttt{buckets} with the result
hash as key and bucket as value.

\paragraph{Largest Group Selection Heuristic}

The heuristic method employed by the voting algorithm is to steer the pipeline
to agree with itself – if multiple generation attempts yielded the same result,
these generation attempts are more likely than outliers.

\paragraph{Error Handling and Fallback}

Queries that voting function fails with execution are that contains in an
\texttt{errors} hash set to provide than being included the calling side all
candidates fail execution, the voting function fails with an error that
contains all collected execution errors to provide diagnostic information for
the calling side.

\paragraph{Limitations and Trade-offs}

The result-based partitioning approach has limitations with subtle errors that
differ by few rows, as these would be treated as completely different result
groups. Furthermore executing every query candidate against the real database
comes along with a significant performance penalty for expensive queries where
the voting step might incur load onto the database, use significant amounts of
memory for loading in all data into memory and slow down the voting as the
entire result needs to be hashed. This system lacks model uncertainty
calibration for SQL task confidence scoring which could be used as one
dimension for partitioning.

Another more advanced optimization could be to group by result-schema and row
count, which still needs to execute the candidate queries partially but does
not load the actual data from the database into memory. For the scope of this
thesis and small real-world application scenarios the cost of this stage is
negligible compared to the llm inference done in previous steps.

The voting mechanism represents the last step of the pipeline function 
and selects the final consensus from all generated and refined candidates, 
and ultimately returns the output of the \textsc{Natural} system.
